C7: Fast software transactional memory for dynamic languages

### 2014-05-15 talk/dls2014/report/report.tex
some tweaks and several notes

\newcommand{\mynote}[2]{%
  \textcolor{red}{%
    \fbox{\bfseries\sffamily\scriptsize#1}%
    {\small$\blacktriangleright$\textsf{\emph{#2}}$\blacktriangleleft$}%
  }%
}
\newcommand\remi[1]{\mynote{Remi}{#1}}

@@

@@

If we start multiple such transactions in multiple threads, the TM
system guarantees that the outcome of running the transactions is
\emph{serializable}. Meaning, the outcome is equal to some sequential
execution of these transactions. Overall, this is exactly what a
single global lock guarantees while still allowing the TM system to
run transactions in parallel as an optimization.

@@

@@

It is well known that using locks to synchronize multiple threads is
hard. They are non-composable, have overhead, may deadlock, limit
scalability, and overall add a lot of complexity. For a better
parallel programming model for dynamic languages, we want to add
another, well-known synchronization mechanism: \emph{atomic blocks}.

@@

@@

@@

To get references to objects that are valid in all threads, we will
use the object's offset inside the segment. Since all segments are
copies of each other, the \emph{Segment Offset (SO)} will point to the
private version of an object in all threads/segments. To then
translate this SO to a real virtual memory address when used inside a
thread, we need to add the thread's segment start address to the
SO. The result of this operation is called a \emph{Linear Address
  (LA)}. This is illustrated in Figure \ref{fig:Segment-Addressing}.

@@

The job of barriers is to ensure complete isolation between transactions
and to register the objects in the read or write set. We insert read
and write barriers before reading or modifying an object except if
we statically know an object to be readable or writable already.
\begin{description}
\item [{Read~Barrier:}] Adds the object to the read set of the current
  transaction. Since our two-step address translation automatically
  resolves the reference to the private version of the object on every
  access anyway, this is not the job of the read barrier anymore.
\item [{Write~Barrier:}] Adds the object to the read and write set of
  the current transaction and checks if all pages of the object are
  private, doing COW otherwise.\\
  Furthermore, we currently allow only one transaction modifying an
  object at a time. To ensure this, we acquire a write lock on the object
  and also eagerly check for a write-write conflict at this point. If
  there is a conflict, we do some contention management to decide which
  transaction has to wait or abort. Eagerly detecting this kind of conflict
  is not inherent to our system, future experiments may show that we
  want to lift this restriction.
\end{description}

@@

To provide atomicity for a transaction, we want to make changes
visible on commit. We also need to be able to completely abort a
transaction without a trace, like it never happened.
\begin{description}
\item [{Commit:}] If a transaction commits, we synchronize all threads
  so that all of them are waiting in a safe point. In the committing
  transaction, we go through all objects in the write set and check if
  another transaction in a different segment read the same object.
  Conflicts are resolved again by either the committing or the other
  transaction waiting or aborting.\\
  We then push all changes of modified objects in private pages to all
  the pages in other segments, including the sharing-segment (segment
  0).
\item [{Abort:}] On abort the transaction will forget about all the
  changes it has done. All objects in the write set are reset by
  copying their previous version from the sharing-segment into the
  private pages of the aborting transaction.
\end{description}

@@

However, the layout of a segment is not uniform and we actually
privatize a few areas again right away. These areas are illustrated in
Figure \ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if SO is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to allocate in the nursery.
\end{description}

@@

Therefore, a thread may be assigned to different segments each time it
starts a transaction. Although, we try to assign it the same segment
again if possible. And a maximum of $N$ transactions may run in
parallel.

@@

@@

The point of the read barrier is to add the object to the read set of
the transaction. This information is needed to detect conflicts
between transactions. Usually, it also resolves an object reference to
a private copy, but since the CPU performs our address translation on
every object access efficiently, we do not need to do that in our
barrier.

@@

Committing a transaction needs a bit more work. First, we synchronize
all threads so that the committing one is the only one running and all
the others are waiting in a safe point. We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads/segments. If we detect a
read-write conflict, we do contention management to either abort us or
the other transaction, or to simply wait a bit.

>>>>

\newcommand{\mynote}[2]{%
  \textcolor{red}{%
    \fbox{\bfseries\sffamily\scriptsize#1}%
    {\small$\blacktriangleright$\textsf{\emph{#2}}$\blacktriangleleft$}%
  }%
}
\newcommand\remi[1]{\mynote{Remi}{#1}}
\newcommand\cfbolz[1]{\mynote{cfbolz}{#1}}

@@

\cfbolz{the above is good, here is something missing: problems with current STM approaches, outlining the intuition behind the new one}

@@

If we start multiple such transactions in multiple threads, the TM
system guarantees that the outcome of running the transactions is
\emph{serializable}. Meaning, the outcome is equal to some sequential
execution of these transactions. This means that the approach provides the same
semantics as using the GIL
while still allowing the TM system to
run transactions in parallel as an optimization.

@@

\cfbolz{a pypy introduction needs to go somewhere, a paragraph or so. maybe in the evaluation section}

@@

 cfbolz{citation again needed for the whole subsection}

It is well known that using locks to synchronize multiple threads is
hard. They are non-composable, have overhead, may deadlock, limit
scalability, and overall add a lot of complexity. For a better
parallel programming model for dynamic languages, we want to implement
another, well-known synchronization mechanism: \emph{atomic blocks}.

@@

\cfbolz{there is an overview paragraph of the idea missing, maybe in the introduction}

\cfbolz{this all feels very much dumping details, needs more overview. why is this info important? the subsubsections don't have any connections}

@@

\cfbolz{this paragraph is hard to understand without giving an example (eg console printing) when it is useful}

@@

To get references to objects that are valid in all threads, we will
use \cfbolz{use for what?} the object's offset inside the segment. Since all segments are
copies of each other, the \emph{Segment Offset (SO)} will point to the
private version of an object in all threads/segments. To then
translate this SO to a real virtual memory address when used inside a
thread, we need to add the thread's segment start address to the
SO. The result of this operation is called a \emph{Linear Address
  (LA)}. This is illustrated in Figure \ref{fig:Segment-Addressing}.

\cfbolz{here it needs to say that this is x86 specific}

@@

The job of barriers is to ensure complete isolation between transactions
and to register the objects in the read or write set. We insert read
and write barriers before reading or modifying an object except if
we statically know an object to be readable or writable already.
\begin{description}
\item [{Read~Barrier:}] Adds the object to the read set of the current
  transaction. Since our two-step address translation automatically
  resolves the reference to the private version of the object on every
  access anyway, the read barrier does not need to do address translation anymore.
\item [{Write~Barrier:}] Adds the object to the read and write set of
  the current transaction and checks if all pages of the object are
  private, doing COW otherwise.\\
  Furthermore, we currently allow only one transaction modifying an
  object at a time. To ensure this, we acquire a write lock on the object
  and also eagerly check for a write-write conflict at this point. If
  there is a conflict, we do some contention management to decide which
  transaction has to wait or abort. Eagerly detecting this kind of conflict
  is not inherent to our system, future experiments may show that we
  want to lift this restriction.
\end{description}

@@

To provide atomicity for a transaction, we want to make changes globally
visible on commit. We also need to be able to completely abort a
transaction without a trace, like it never happened.
\begin{description}
\item [{Commit:}] If a transaction commits, we synchronize all threads
  so that all of them are waiting in a safe point. In the committing
  transaction, we go through all objects in the write set and check if
  another transaction in a different segment read the same object.
  Conflicts are resolved again by either the committing or the other
  transaction waiting or aborting.\\
  We then push all changes of modified objects in private pages to all
  the pages in other segments, including the sharing-segment (segment
  0). \cfbolz{can it really happen that you push pages to other segments? I thought it's always just back to the sharing segment}
\item [{Abort:}] On abort the transaction will forget about all the
  changes it has done. All objects in the write set are reset by
  copying their previous version from the sharing-segment into the
  private pages of the aborting transaction.
  \cfbolz{why doing any copying? aren't the pages re-shared instead?}
\end{description}

\cfbolz{random question: did we investigate the extra memory requirements? we should characterize memory overhead somewhere, eg at least one byte per object for the read markers}

@@

However, the layout of a segment is not uniform and we actually
privatize a few areas again right away. These areas are illustrated in
Figure \ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if SO is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

Therefore, a thread may be assigned to different segments each time it
starts a transaction. Although, we try to assign it the same segment
again if possible.

@@

\cfbolz{maybe use "young" and "old" generation, if there are only two}

@@

The point of the read barrier is to add the object to the read set of
the transaction. This information is needed to detect conflicts
between transactions. In other STM systems, it also resolves an object reference to
a private copy, but since the CPU performs our address translation on
every object access efficiently, we do not need to do that in our
barrier.

@@

Committing a transaction needs a bit more work. First, we synchronize
all threads so that the committing one is the only one running and all
the others are waiting in a safe point. We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads/segments. If we detect a
read-write conflict, we do contention management to either abort us or
the other transaction, or to simply wait a bit. \cfbolz{why does waiting help?}
